Architectures for workflow integration in science gateways

### 2016-04-20 paper.tex
Changed labels to Roman numbers on subfigures of Figure 2

\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
\usepackage{subfig}
\usepackage{svg}
\usepackage{url}
\usepackage{flushend}
\usepackage{xspace}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{multirow}

@@

@@

\begin{abstract}
%% Text of abstract

\end{abstract}

@@

See Figure~\ref{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored. In this model, task and data control are both
initiated from the science gateway. Interactions \texttt{b} and
\texttt{c$_2$} are initiated from the workflow engine while
\texttt{c$_1$} comes from other parts of the science gateway, for
instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}),
CIPRES~\cite{miller2010creating} and LONI Pipeline
Environment~\cite{dinov2009efficient}.
\todo{Add description of a real system.}

@@

See Figure~\ref{archi:service}. The workflow engine is available
externally to the science gateway, in a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g. RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
tasks, or as a white box that allows for some interaction with
them. The workflow engine is responsible for controlling the tasks on
the infrastructure (\texttt{b}) and for performing the required data
transfers to execute them (\texttt{c$_2$}). User data is usually
managed through the science gateway (\texttt{c$_1$}), although it
might as well be delivered by the workflow engine directly to the
user.

@@

See Figure~\ref{archi:sub-task}. The workflow engine is a particular
task that can submit sub-tasks to the science gateway through
interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks but their execution is delegated to
the science gateway that executes them on the infrastructure through
interaction \texttt{b}. Although the science gateway has no global
vision of the workflow, it can keep track of the sub-tasks submitted
by a given task, for instance to be able to cancel them when the task
is canceled. The science gateway may also implement mechanisms to
facilitate the handling of task dependencies, for instance basic
dependency lists as available in most batch managers (see for instance
attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

\subsection{Pool model}
\label{sec:pool}
See Figure~\ref{archi:agent}. Workflows are submitted by the science
gateway to a pool through interaction \texttt{d}. Agents connect to
the pool asynchronously to retrieve and execute workflows through
interaction \texttt{f}. Agents may be started according to various
policies, for instance to ensure load balancing. Agents may wrap
different types of workflow engines. Workflow engine controls tasks
and data on the infrastructure through interactions \texttt{b} and
\texttt{c$_2$}, science gateway transfers user data through
interaction \texttt{c$_1$}, and administrator installs workflows
through interaction \texttt{a$_1$} and \texttt{a$_2$}.

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}, see
Figure~\ref{fig:shiwa-pool-architecture}. In Figure
~\ref{fig:shiwa-pool-architecture}, interaction \texttt{d} of
Figure~\ref{archi:agent} is implemented in 3 different types of calls
to the pool: workflow submission (\texttt{1} \& \texttt{2}), workflow
status retrieval (\texttt{a} and \texttt{b}), and workflow retrieval
(\texttt{13} \& \texttt{14}). In Figure
~\ref{fig:shiwa-pool-architecture}, interaction \texttt{f} of
Figure~\ref{archi:agent} is implemented through 2 types of calls:
workflow instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}),
and workflow instance update (\texttt{11} and \texttt{12}). Workflow
instance retrieval is used by the agents to fetch work from the
pool. Workflow instance update is used by the agents to update
workflow statuses. In the SHIWA pool, agents can wrap different types
of workflow engines to execute workflows expressed with different
languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and \texttt{9} on
Figure~\ref{fig:shiwa-pool-architecture} are used by workflow engine
plugins to declare their supported language and launch engines, and by
workflow engines to report their status to the agent. These calls are
specific to the SHIWA Pool implementation of the \texttt{agent}
component and therefore have no corresponding representation in
Figure~\ref{archi:agent}.

@@

See Figure~\ref{archi:nested}. In nested workflows
(Figure~\ref{archi:nested}-Left), a task of a \emph{parent} workflow
executed by the \emph{parent} engine is itself a workflow, called
\emph{child} workflow, that is executed by a \emph{child} engine. The
parent and child engines might use different workflow description
languages. They might also execute workflows on different
infrastructures. The parent workflow is also called meta-workflow. The
science gateway communicates with the parent engine through
interaction \texttt{*$_1$}. The science gateway also communicate with
the infrastructure to transfer user data through abstract interactions
\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
with the infrastructure through abstract interactions as well,
\texttt{*$_4$} and \texttt{*$_5$}. The parent engine communicates with
the child engine through abstract interaction
\texttt{*$_6$}. Administrator installs workflows in the science
gateway through interaction \texttt{a$_2$}, and installs software on
infrastructures through interactions \texttt{a$_1$}.

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model (see
Figure~\ref{archi:nested}-Right) as this is the most used
architecture. In such an instantiation, we assume that the parent and
child workflow engines are distinct pieces of software that require
different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case, then workflow services can be
collapsed in a single one with a \texttt{d} interaction with
itself. Workflow engines communicate with infrastructures using
\texttt{b} and \texttt{c$_2$}. Science gateway transfer user data to
infrastructures using \texttt{c$_1$} interactions.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Science Gateway for nested workflow execution with
service invocation as represented in Figure~\ref{archi:nested}. The
parent workflow engine is \texttt{WS-PGRADE}, invoked as a service in
the Science Gateway (step \texttt{1} on
Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
Figure~\ref{archi:nested}). Ten different child engines can be used by
nested workflows, invoked through the \texttt{Submission service}
(step \texttt{2}, interaction \texttt{d}). Each of these engines can
submit jobs and transfer data to a distributed computing
infrastructure (\texttt{DCI}, step \texttt{3}, interactions \texttt{b}
and \texttt{c$_2$}). Data interactions (\texttt{c}) and application
porting ones (\texttt{a}) are not represented on
Figure~\ref{fig:shiwa-architecture}.

@@

See Figure~\ref{archi:import}. This is an abstract model that we
instantiate with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. This was implemented in the SHIWA Science Gateway through
the IWIR language that provided a common language for portability
across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013}.

>>>>

\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
\usepackage[subrefformat=parens,labelformat=parens]{subfig}
\usepackage{svg}
\usepackage{url}
\usepackage{flushend}
\usepackage{xspace}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{multirow}

@@

\renewcommand\thesubfigure{\Roman{subfigure}}

@@

\begin{abstract}
%% Text of abstract
  We investigate software architectures to integrate workflow engines
  in science gateways. Based on our experience with the development
  and operations of the CBRAIN and VIP science gateways, and our past
  involvement in the SHIWA project, we abstract the software
  components and interactions in 6 architectures that we compare using
  metrics measuring integration effort, robustness, extensibility,
  scalability and other specific features. \todo{a word on results}
\end{abstract}

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored. In this model, task and data control are both
initiated from the science gateway. Interactions \texttt{b} and
\texttt{c$_2$} are initiated from the workflow engine while
\texttt{c$_1$} comes from other parts of the science gateway, for
instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}),
CIPRES~\cite{miller2010creating} and LONI Pipeline
Environment~\cite{dinov2009efficient}.
\todo{Add description of a real system.}

@@

See Figure~\subref*{archi:service}. The workflow engine is available
externally to the science gateway, in a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g. RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
tasks, or as a white box that allows for some interaction with
them. The workflow engine is responsible for controlling the tasks on
the infrastructure (\texttt{b}) and for performing the required data
transfers to execute them (\texttt{c$_2$}). User data is usually
managed through the science gateway (\texttt{c$_1$}), although it
might as well be delivered by the workflow engine directly to the
user.

@@

See Figure~\subref*{archi:sub-task}. The workflow engine is a particular
task that can submit sub-tasks to the science gateway through
interaction \texttt{e}. The workflow engine keeps track of the
dependencies between the sub-tasks but their execution is delegated to
the science gateway that executes them on the infrastructure through
interaction \texttt{b}. Although the science gateway has no global
vision of the workflow, it can keep track of the sub-tasks submitted
by a given task, for instance to be able to cancel them when the task
is canceled. The science gateway may also implement mechanisms to
facilitate the handling of task dependencies, for instance basic
dependency lists as available in most batch managers (see for instance
attribute \texttt{depend} of option \texttt{-W} in
Torque\footnote{\url{http://docs.adaptivecomputing.com}}).

@@

\subsection{Pool model}
\label{sec:pool}
See Figure~\subref*{archi:agent}. Workflows are submitted by the science
gateway to a pool through interaction \texttt{d}. Agents connect to
the pool asynchronously to retrieve and execute workflows through
interaction \texttt{f}. Agents may be started according to various
policies, for instance to ensure load balancing. Agents may wrap
different types of workflow engines. Workflow engine controls tasks
and data on the infrastructure through interactions \texttt{b} and
\texttt{c$_2$}, science gateway transfers user data through
interaction \texttt{c$_1$}, and administrator installs workflows
through interaction \texttt{a$_1$} and \texttt{a$_2$}.

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}, see
Figure~\ref{fig:shiwa-pool-architecture}. In Figure
~\ref{fig:shiwa-pool-architecture}, interaction \texttt{d} of
Figure~\subref*{archi:agent} is implemented in 3 different types of calls
to the pool: workflow submission (\texttt{1} \& \texttt{2}), workflow
status retrieval (\texttt{a} and \texttt{b}), and workflow retrieval
(\texttt{13} \& \texttt{14}). In Figure
~\ref{fig:shiwa-pool-architecture}, interaction \texttt{f} of
Figure~\subref*{archi:agent} is implemented through 2 types of calls:
workflow instance retrieval (\texttt{4}, \texttt{5} and \texttt{6}),
and workflow instance update (\texttt{11} and \texttt{12}). Workflow
instance retrieval is used by the agents to fetch work from the
pool. Workflow instance update is used by the agents to update
workflow statuses. In the SHIWA pool, agents can wrap different types
of workflow engines to execute workflows expressed with different
languages. Calls \texttt{0}, \texttt{7}, \texttt{8} and \texttt{9} on
Figure~\ref{fig:shiwa-pool-architecture} are used by workflow engine
plugins to declare their supported language and launch engines, and by
workflow engines to report their status to the agent. These calls are
specific to the SHIWA Pool implementation of the \texttt{agent}
component and therefore have no corresponding representation in
Figure~\subref*{archi:agent}.

@@

See Figure~\subref*{archi:nested}. In nested workflows
(Figure~\subref*{archi:nested}-Left), a task of a \emph{parent} workflow
executed by the \emph{parent} engine is itself a workflow, called
\emph{child} workflow, that is executed by a \emph{child} engine. The
parent and child engines might use different workflow description
languages. They might also execute workflows on different
infrastructures. The parent workflow is also called meta-workflow. The
science gateway communicates with the parent engine through
interaction \texttt{*$_1$}. The science gateway also communicate with
the infrastructure to transfer user data through abstract interactions
\texttt{*$_2$} and \texttt{*$_3$}. Both workflow engines communicate
with the infrastructure through abstract interactions as well,
\texttt{*$_4$} and \texttt{*$_5$}. The parent engine communicates with
the child engine through abstract interaction
\texttt{*$_6$}. Administrator installs workflows in the science
gateway through interaction \texttt{a$_2$}, and installs software on
infrastructures through interactions \texttt{a$_1$}.

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model (see
Figure~\subref*{archi:nested}-Right) as this is the most used
architecture. In such an instantiation, we assume that the parent and
child workflow engines are distinct pieces of software that require
different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case, then workflow services can be
collapsed in a single one with a \texttt{d} interaction with
itself. Workflow engines communicate with infrastructures using
\texttt{b} and \texttt{c$_2$}. Science gateway transfer user data to
infrastructures using \texttt{c$_1$} interactions.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Science Gateway for nested workflow execution with
service invocation as represented in Figure~\subref*{archi:nested}. The
parent workflow engine is \texttt{WS-PGRADE}, invoked as a service in
the Science Gateway (step \texttt{1} on
Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
Figure~\subref*{archi:nested}). Ten different child engines can be used by
nested workflows, invoked through the \texttt{Submission service}
(step \texttt{2}, interaction \texttt{d}). Each of these engines can
submit jobs and transfer data to a distributed computing
infrastructure (\texttt{DCI}, step \texttt{3}, interactions \texttt{b}
and \texttt{c$_2$}). Data interactions (\texttt{c}) and application
porting ones (\texttt{a}) are not represented on
Figure~\ref{fig:shiwa-architecture}.

@@

See Figure~\subref*{archi:import}. This is an abstract model that we
instantiate with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. This was implemented in the SHIWA Science Gateway through
the IWIR language that provided a common language for portability
across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013}.
